Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-12-19 Manuscript/manuscript.tex

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89_POPC-struct.pdf}
  %\includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-POPC models compared with experiments~\cite{kucerka11} at 303~K. \\
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-POPC models
    compared with experiments \cite{ferreira13} at 300~K.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Open/closed symbols are used for palmitoyl/oleoyl chains of POPC. \\
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
  } 
\end{figure}

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of head group order parameters and P-N vector orientation as a function of
    a molar fraction of the cationic surfactant dihexadecyldimethylammonium in a POPC bilayer
    from simulations and experiments \cite{scherer89} at 313 K.
  }
  \todo{Joe: currenct figure depicts simulations with OPC3 water. Simulations with SPCE are running now.}
\end{figure}

@@

=======
%% ****** Start of file aiptemplate.tex ****** %
%%
%%   This file is part of the files in the distribution of AIP substyles for REVTeX4.
%%   Version 4.1 of 9 October 2009.
%%
%
% This is a template for producing documents for use with 
% the REVTEX 4.1 document class and the AIP substyles.
% 
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.

%\documentclass[aip,graphicx]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

%\usepackage{graphicx}

%\draft % marks overfull lines with a black rule on the right
%\documentclass[pre,aps,floatfix,authordate1-4,twocolumn]{revtex4-1}
%\documentclass[pre,aps,floatfix,authordate1-4]{revtex4-1}

\documentclass[aip,jcp,twocolumn]{revtex4}
%\documentclass[aip,jcp]{revtex4}
%\documentclass{article}

%\documentclass[aps,prl,preprint,groupedaddress]{revtex4}

\begin{document}

% Use the \preprint command to place your local institutional report number 
% on the title page in preprint mode.
% Multiple \preprint commands are allowed.
%\preprint{}

\title{Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization} %Title of paper

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
% Explanatory text should go in the []'s, 
% actual e-mail address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro for the type of information

% \affiliation command applies to all authors since the last \affiliation command. 
% The \affiliation command should follow the other information.

\author{Josef Melcr}
\author{Hector Martinez-Seara}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Ji{\v r}{\' i} Kolafa}
\affiliation{Department of Physical Chemistry, Institute of Chemical Technology, Prague 6, Czech Republic}
\author{Pavel Jungwirth}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Department of Physics, Tampere University of Technology, P.O. Box 692, FI-33101
Tampere, Finland}

\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}

% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass). 
% \noaffiliation is required (may also be used with the \author command).
%\collaboration{}
%\noaffiliation

\date{\today}

\begin{abstract}
  % insert abstract here
Binding affinities and stoichiometries of Na$^+$ and Ca$^{2+}$ ions to phospholipid
bilayers are of paramount significance in the properties and functionality of cellular
membranes. Current estimates of binding affinities and
stoichiometries of cations are, however, controversial due to limitations in the available experimental and
theoretical methods. Experimentally one can assess these parameters by titrating
several membrane properties as a function of the ion concentration. However, the
interpretation of the experiments relies on theoretical models, which give
the best conception of the ion binding as direct experiments are not available. 
Classical molecular dynamics (MD) simulations provide details of the ion binding process with atomistic
resolution, therefore offering all the necessary information to interpret
experimental data without the need to resort to simplified models. However, the accuracy
of the available lipid models when interacting with ions is not sufficient for
such interpretation. In this work, we improve the binding details of Na$^+$ and Ca$^{2+}$ ions to the
1-Palmitoyl-2-oleoyl-phosphatidylcholine (POPC) bilayer by implicitly including electronic
polarization as a mean field correction, known as the electronic continuum correction (ECC).
This is applied to the partial charges of a selected state of the art POPC lipid model for MD simulations.
Our improved ECC-POPC model reproduces not only the
experimentally measured structural parameters for the ion-free membrane, but also the
response of lipid headgroup to a bound positive charge, and the binding affinities of 
Na$^+$ and Ca$^{2+}$ ions. With our new model we also observe negligible binding of Na$^+$ ions to POPC
bilayer and stronger interactions of Ca$^{2+}$ primarily with phosphate oxygens,
in agreement with the previous interpretations of the experimental spectroscopic data.
The new model results in Ca$^{2+}$ ions forming complexes with one to three POPC molecules with almost
equal probabilities, suggesting more complex binding stoichiometry than simple models
previously used to interpret the NMR data.
The results of this work pave the way to quantitative MD simulations of complex
biochemical systems with realistic electrostatic interactions in the vicinity of cellular
membranes.
\end{abstract}

%\pacs{}% insert suggested PACS numbers in braces on next line

\maketitle %\maketitle must follow title, authors, abstract and \pacs

% Body of paper goes here. Use proper sectioning commands. 
% References should be done using the \cite, \ref, and \label commands

\section{Introduction}
Ion interactions with cellular membranes play a key role in critical biological processes~\cite{seelig90,cevc90}. Ions, especially multivalent cations, modify general properties of the membrane which ultimately modulate their embedded transmembrane proteins~\cite{cevc90,tocanne90, pabst07}. Ions also have a more direct effect by modulating lipid--, protein--, and sugar--lipid interactions at the membrane surface. For example, Ca$^{2+}$  is crucial in neural signal propagation. It promotes membrane fusion by bridging lipids in the vesicles carrying neurotransmitters and the neuron synapsis membrane~\cite{REF}. Ca$^{2+}$ also participates in the T-cell receptor activation. It induces the detachment of the positively charged cytosolic tails of the CD3 protein complex from the negatively charged intracellular membrane making it accessible for Lck protein~\cite{Shi2013Ca2}. A final example involving sugar-lipid interactions mediated by ions is that Ca$^{2+}$ can modulate the presentation of the sugars found in the PI(4,5)P$_2$ lipid which ultimately modulates phospholipase C delta 1 pleckstrin homology domain (PLC $\delta$1-H)~\cite{Bilkova2017Calcium}. Interestingly, this modulation is lost when using Mg$^{2+}$ instead of Ca$^{2+}$ illustrating the selectivity of these processes towards particular ions. Despite our increasing understanding of the role of ions in cell membrane-related processes the exact molecular details of the mechanisms behind such processes remain elusive. \todo{Hector: I left one unknown citation for Pavel.}

Direct measurements of ion-membrane interactions in real biological systems are difficult. 
Hence, simplified lipid bilayers are often used as entry-level models to shed light on the role of ions in complex biological membranes~\cite{scherer87,seelig90,cevc90}. 
For this reason, interactions of biologically relevant cations, especially Na$^+$ and Ca$^{2+}$, with zwitterionic phosphocholine (PC) bilayers have been widely studied in experiments~\cite{akutsu81, altenbach84, seelig90, cevc90, tocanne90, binder02, pabst07, uhrikova08} and classical MD simulations~\cite{bockmann03, bockmann04, Berkowitz12, melcrova16, javanainen17}. The details of ion binding are, however, not fully consistent in the literature. Non-invasive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering, and infrared spectroscopies mainly suggest that Na$^+$ ions exhibit negligible binding to PC lipid bilayers. On the contrary, Ca$^{2+}$ is observed to specifically bind to a couple of PC molecules using their phosphate groups~\cite{hauser76, hauser78, herbette84, akutsu81, altenbach84, binder02, pabst07, uhrikova08}. On the other hand, most atomistic resolution molecular dynamics (MD) simulation models predict a stronger binding for the cations~\cite{catte16}. For example, simulations report Na$^+$ accumulating in the lipid interface~\cite{bockmann03} or Ca$^{2+}$ binding up to 4 PC lipids simultaneously, including not only interactions with phosphates but also with carbonyl oxygens~\cite{bockmann04, melcrova16, javanainen17}.
%{\color{red} Also, several experiments have also been interpreted to support the predictions from earlier MD simulations~\cite{bockmann03, vacha09a}. \todo{remove? Why we need this?}}

Recent studies within the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16} made an attempt to resolve these apparent controversies. A direct comparison of ion binding affinities to PC bilayers between simulations and experiments was performed using the electrometer concept~\cite{seelig87}. Namely, the changes in NMR order parameters of the headgroup upon addition of ions is directly compared to the MD simulations results. 
Analyzing massive amounts of data collected by an open collaboration method, 
it was concluded that the accuracy of the current state of the art lipid models for MD simulations is not sufficient for a detailed interpretation of the interactions of cations with PC lipid bilayers~\cite{catte16}.

In this work, we improve the cation binding in a POPC bilayer in MD simulations 
via the implicit inclusion of the electronic polarizability in the polar region of phospholipids. 
For this, we apply an electronic continuum correction (ECC)~\cite{leontyev11}, 
in which scaling of the atomic partial charges is used to account for the missing electronic polarizability in MD simulations. 
Such an approach has been previously shown to improve the behavior of ions in bulk water simulations ~\cite{martinek17, Pluharova2014, kohagen14, kohagen16}. As a starting template for our POPC model, we will use the Lipid14~\cite{dickson14}. This force field provides one of the best available descriptions of cation binding \cite{catte16}. 
The newly developed ECC-POPC model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the accuracy comparable to the best state of the art lipid models, while significantly improving the membrane binding affinities to sodium and calcium cations.

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in standard MD simulation force fields has been considered a serious issue since the early days of lipid bilayer simulations. In this work, we circumvent the demanding explicit inclusion of electronic polarization effects~\cite{lucas12, chowdhary13} by including the electronic part of polarizability in lipid bilayer simulations implicitly via the electronic continuum correction (ECC)~\cite{leontyev11}. Technically, this is similar to the phenomenological charge-scaling applied in earlier studies of surfactants, lipids or ionic liquids \cite{jonsson86, egberts94, beichel14}. However, the present concept of ECC is physically well justified and has
obtained theoretical support~\cite{leontyev09, leontyev10, leontyev11, leontyev14}.

According to ECC, the electronic polarizability can be modeled in non-polarizable MD in a mean-field way by 
embedding the ions in a homogeneous dielectric continuum with a dielectric constant $\epsilon_{el}$, which is the electronic part of the dielectric constant of the medium~\cite{leontyev11}. Following Coulomb's law,  ECC can be directly incorporated by scaling the charges with a constant scaling factor of $f_q = \epsilon _{el} ^{-1/2}$, giving
\begin{equation}
Q^{ECC} = f_q \cdot Q
\end{equation}
for the ECC corrected charges in simulation. 
Given that the  high frequency dielectric constant of water is $\epsilon _{el} = 1.78$ (\textit{i.e.,} the square of the refraction index), the scaling factor for ions in water is roughly $f_q \approx 0.75$. This scaling factor successfully improves the accuracy of simulations of solvated ions, when quantitatively compared with neutron scattering data \cite{kohagen14,kohagen16, Pluharova2014, martinek17}.
It is important to note that the 
high frequency dielectric constant 
is around 2 for almost any biologically relevant environment \cite{leontyev11}.
The dielectric discontinuity in a lipid bilayer arises
from the orientational polarization of the molecules. 
Thus, the same correction for the electronic polarizability can be 
applied throughout the lipid bilayer interface.

While using a scaling factor of $f_q = 0.75$ for ions in water is well justified in the ECC theory \cite{leontyev11}, it is not clear whether the same factor should be applied to partial charges used to describe molecules in MD models, e.g., lipids in our case. Unlike the total charge of an ion, atomic partial charges within molecules are not physical observables. Several schemes exist for the assignment of partial charges for biomolecules~\cite{Hu2007}, where the restrained electrostatic potential method (RESP) is the most commonly used~\cite{RESP_paper, Singh1984}. Considering that water is often included in RESP calculations or charges are refined to improve certain experimental observables, the electronic polarizability effect of the solvent is included in modern force fields to some extent~\cite{RESP_paper, Singh1984, jorgensen96, ipolq2013, benavides17}. Thus, the application of the ECC scaling factor, $f_q$, to the existing partial charges in molecules does not universally follow $f_q = \epsilon _{el} ^{-1/2}$. Instead, a consistent scaling factor would lie between the value of 0.75  (\textit{i.e.,} no electronic polarizability included in the original partial charges) and 1 (\textit{i.e.,} electronic polarizability fully included in the original partial charges).

Here we develop ECC-POPC lipid model that accurately describes the binding 
of sodium and calcium ions to the lipid bilayer. 
The Lipid14~\cite{dickson14} force field parameters 
(available in a Gromacs format from Ref. \citenum{lipid14files}) were used as a starting 
point. 
This model provides the best response of the head group to ions among the available 
lipid models (see Figs.~2~and~5 in Ref.~\citenum{catte16}). Additionally, the Lipid14 model 
has a relatively realistic head group, glycerol backbone, and acyl chain structure~\cite{dickson14,botan15}.
We applied the ECC correction 
to the Lipid14 model of POPC by scaling of 
partial charges of the head group, glycerol 
backbone, and carbonyl regions. 
These are the polar parts of phospholipids which can 
contribute to the cation binding.

To reproduce the experimental ion binding affinities,
we scanned the optimal interval for the scaling factor, $f_q~\in~(0.75, 1.0)$.
The ion binding affinity was evaluated against experiments
using the head group order parameters and the electrometer concept \cite{seelig87,catte16},
as discussed more detail in the next section.
Scaling down the partial charges reduced the ion binding affinity.
We found an optimal binding affinity with the scaling factor of $f_q = 0.8$,
which is only slightly higher than the theoretical one ($f_q=0.75$).
Directly applying
the 0.8 scaling to the partial charges of the head group, the glycerol backbone, and
the carbonyls reduced the area per lipid to 60~\AA$^2$. This area is smaller than in the
original Lipid14 model ($65.6 \pm 0.5$~\AA$^2$)~\cite{dickson14} and in experiments
(64.3~\AA$^2$) \cite{kucerka11}. The decrease of the area per lipid arises from a
reduced hydration of the lipid head group region after scaling charges which effectively
reduces head group polarity. We solve this difficulty by reducing the effective radii of
the modified atoms by lowering the $\sigma$ parameters in the Lennard-Jones potential by a
factor of $f_\sigma = 0.89$. The same solution was taken for the ECC-ions in aqueous
solvent \cite{kohagen14, kohagen16, Pluharova2014, martinek17}. After reducing the $\sigma$ parameters,
the area per molecule is restored to the experimental value (Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Comparing MD simulation to NMR experiments, we can validate the ion
binding affinity in lipid bilayer simulations using the ''electrometer concept''~ \cite{seelig87, catte16}.
This method relies on the experimental observation that the C-H bond order parameters of $\alpha$ and $\beta$ carbons in a PC lipid head group (Fig.~\ref{simVSexpNOions}) are proportional to the amount of charge bound per lipid~\cite{seelig87}.
%Comparing this response is an excellent way to validate force fields as both NMR and MD can accurately determine order parameters ~\cite{catte16, ollila16}.
The order parameters for all C-H bonds in lipid molecules can be accurately measured using $^2$H NMR or $^{13}$C NMR techniques \cite{ollila16}. From MD simulations the order parameters can be calculated using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C-H bond and membrane normal. Angular brackets point to the average over all sampled configurations.

The relation between the amount of the bound charge per lipid, $X^\pm$, and the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$, is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx m_i \frac{4 }{3\chi}X^\pm,
\end{equation}
where $i$ refers to either $\alpha$ or $\beta$ carbon, $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge, $\chi$ is the quadrupole coupling constant ($\chi \approx$\,167\,kHz), and $m_i$ is an empirical constant depending on the valency and position of the bound charge.

The measured change of the order parameter depends on the head group response to the bound charge and on the amount of the bound charge (\textit{i.e.,} $m_i$ and $X^\pm$ in Eq.~\ref{OPchangeEQ}, respectively).  The empirical factor $m_i$ has to be adequately quantified before the electrometer concept can be used to analyze the binding affinities. This calibration has been done experimentally for a wide range of systems~\cite{seelig87, beschiasvili91}. To calibrate the response of the head group order parameter to the bound charge in simulations, we use experimental data for a strong cationic surfactant dihexadecyldimethylammoniumbromide  (DHAB) mixed with a POPC bilayer~\cite{scherer89}. DHAB\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
  -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cationic surfactant having two acyl chains and bearing a unit charge in the hydrophilic end. Thus, it is expected to locate in the bilayer similarly to the phospholipids, and its molar ratio then gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

\subsection{Salt concentrations and binding affinity}
NMR experiments report the used salt concentrations in two different ways when measuring head group order parameters.
Some use the salt concentrations in water before solvating the lipids~\cite{akutsu81}, $C_{ion}'$,
while others use atomic absorption spectroscopy and report the salt concentration in
the supernatant after the solvation of lipids~\cite{altenbach84}, $C_{ion}$. In this work,
%we use the latter definition. In particular,
we measure the latter concentration in the aqueous
bulk region using the farthest point from the lipid bilayer in the aqueous phase.
Notice that the former concentration was used by Catte et al.~\cite{catte16}.
Although these two definitions vary when applied
to Ca$^{2+}$, they do not significantly affect the conclusions of this work.

To quantify the ion binding affinities to a lipid bilayer, we calculate the relative surface excess of ions with respect to water, $\Gamma_{ion}^{water}$~\cite{chattorajBOOK}. Such a quantity compares the adsorption of ions to the adsorption of water molecules at the interface without the necessity of defining a Gibbs dividing plane between the membrane interior and the water bulk region. In our simulations, we only assume that the interface is located between the ion-free hydrophobic interior of the lipid bilayer and the aqueous region far from the membrane.  Such a setup and the above definition of bulk ion concentration provide a simple relation for the relative surface excess $\Gamma_{ion}^{water}$ for simulations of lipid bilayers,
 \begin{equation}\label{surfexcess}
  \Gamma_{ion}^{water}=\frac{1}{2A_b} \left ( n_{ion} - n_{water} \frac{C_{ion}}{C_{water}} \right ).
\end{equation}
Here, $n_{water}$ and $n_{ion}$ are the total numbers of water molecules and ions in the system,
$C_{water}$ and $C_{ion}$ are their respective bulk concentrations in the aqueous phase,
and $A_b$ is the area of the unit cell in the membrane plane.
The total area of the interface is then twice the area of the membrane, i.e., $2A_b$,
because the bilayer has an interface at each of the two leaflets.

\subsection{Validation of lipid bilayer structure against NMR and scattering experiments}
The structures of lipid bilayers in simulations without ions were validated against NMR by calculating the order parameters for the C-H bonds and \mbox{x-ray} scattering experiments by the scattering form factors. NMR order parameters validate the structures sampled by the individual lipid molecules with atomic resolution. The simulated order parameters were calculated for all C-H bonds in lipid molecules from Eq. \ref{OP}. Scattering form factors validate the dimensions of the lipid bilayer (e.g., thickness and area per molecule). Form factors were calculated using

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of a POPC bilayer in pure water or at varying salt concentrations consisted of 128 POPC molecules
and approximately 50~water molecules per lipid in an orthorhombic simulation box with periodic boundary conditions.
The SPC/E~\cite{Berendsen1987} water model was used in all ECC-POPC model simulations reported
in the main text. This model was also used in the previous parametrization
of ECC-ions \cite{martinek17, kohagen16, Pluharova2014} because its lowered dielectric constant is consistent with
the ECC concept~\cite{leontyev11, leontyev14}.
The robustness of our approach with other water models (OPC \cite{Izadi14},
OPC3~\cite{Izadi16}, TIP3P~\cite{jorgensen83}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005})
is tested in the SI.
Sodium, calcium, and chloride ions were modeled as ECC-ions
with parameters from Refs. \citenum{martinek17, Pluharova2014}. Scaled charges and Lennard-Jones radii for atoms
forming the POPC lipid are derived in this work starting from the Lipid14 force field~\cite{dickson14}.
For comparison, simulations with the Lipid14 model~\cite{dickson14} and
ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006} or ECC-ions \cite{martinek17, kohagen16, Pluharova2014}
were also performed. The TIP3p water model~\cite{jorgensen83} was used in all simulations with the original Lipid14 model.
Simulation data for the Lipid14 model with \AA{}qvist ions~\cite{aqvist90} and TIP3P~\cite{jorgensen83} water models were taken directly from
Refs.~\citenum{catte16,lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC,lipid14POPC1000mMNaClfiles}.  
% SAMULI: I putted the trajectory without ions back, because it has been used as well.

MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4). The simulation parameters used
in this work are summarized in Table~S1. 
Compatibility with the openMM simulation package \cite{openmm7} was also
tested in SI. 
The simulated trajectories and parameters files are available at 
Refs.~\citenum{lipid14_POPC_Dang_and_ECC-ions_files, ECC-POPC_nacl_cacl2_files, ECC-POPC_nacl_cacl2_wat-models_files} 
\todo{To be uploaded to Zenodo, references are only placeholders. }.

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
The topology of dihexadecyldimethylammonium was created with the automated topology builder \cite{malde11}. The AmberTools program~\cite{amber} was then used to generate the Amber-type force field parameters. These parameters were then converted to the Gromacs format with the acpype tool \cite{acpype}. The partial charges were then manually modified in the non-scaled version to approximately match similar segments in Lipid14 \cite{dickson14}. For simulation with ECC-POPC, the partial charges of the headgroup were scaled, i.e., $f_q=0.8$, and the $\sigma$ of the scale atoms reduced as in POPC-ECC.

The cationic surfactants were randomly mixed among the phospholipids to form bilayer structures with mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer. All these systems contained 50 POPC molecules per leaflet, 6340 water molecules, and 6, 14, 21, 35, or 50 surfactants per leaflet.  Chloride counter ions were used in simulations, while bromide was the counter ion in the experiment~\cite{scherer89}. For POPC either the Lipid14 with TIP3P~\cite{jorgensen83} or ECC-POPC with SPC/E~\cite{Berendsen1987}  were used. The first 20~ns of the total simulation time of 200~ns was considered as an equilibration time and was omitted from the analysis. Sufficient lipid neighbor exchange occurred during the simulations.
The simulated trajectories and parameters files are available at Ref.~\citenum{ECC-POPC_cationic-surf_files} 
\todo{To be uploaded to Zenodo, references are only placeholders. }.
The trajectories and simulation files for the unscaled systems are available from
Refs.~\citenum{POPClipid14T313K,POPClipid1410perCATsurfT313K,POPClipid1420perCATsurfT313K,POPClipid1430perCATsurfT313K,POPClipid1442perCATsurfT313K,POPClipid1450perCATsurfT313K}.

\subsection{Structural parameters of pure ECC-POPC model membrane are in agreement with experiments}

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89_POPC-struct.pdf}
  %\includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-POPC models compared with experiments~\cite{kucerka11} at 303~K. \\
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-POPC models
    compared with experiments \cite{ferreira13} at 300~K.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$ 
    (Bayesian estimate of 95\% confidence interval \cite{scipy}).
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Open/closed symbols are used for palmitoyl/oleoyl chains of POPC. \\
    Bottom: Chemical structure of POPC and the labeling of hydrocarbon segments.
  } 
\end{figure}

First, we present results for bilayers in pure water.
The ECC-POPC and Lipid14 models both reproduce the experimental x-ray scattering form factors
of a POPC bilayer with a comparable accuracy in Fig.~\ref{simVSexpNOions}.
The area per lipid from the Lipid14 model is $\approx$1\AA{} larger than the
experimental value in Table~\ref{tab:apls}, while the value from the ECC-POPC model
is $\approx$1\AA{} smaller
The values of the area per lipid of the ECC-POPC model vary
when simulated with different water models (i.e., 62.2--66.8 \AA{}, see Table~S2 in SI),
while still being close to the experimentally reported values.
We can thus conclude that the ECC-POPC model reproduces the experimental dimensions of the POPC
lipid bilayer with the comparable accuracy to other state of the art lipid models~\cite{ollila16}.

Similarly, the acyl chain order parameters of the ECC-POPC model, as well as the Lipid14 model~\cite{dickson14}, agree with the experimental values within the error bars, as presented in Fig.~\ref{simVSexpNOions}. Notably, the experimentally measured forking and small order parameter values of $C_2$ segment in {\it sn}-2 chain are well reproduced by both models. This feature has been suggested to indicate that the carbonyl of {\it sn}-2 chain is directed towards the water phase, in contrast to the carbonyl in {\it sn}-1 chain, which would orient more along the bilayer plane~\cite{seelig75,schindler75,gawrisch92}.
While this arrangement may be a relevant feature for the ion binding details, it is not fully reproduced by other available lipid models~\cite{ollila16}.
%This is expected as the ECC model does not modify the already highly optimized parameters of this region in Lipid14

The order parameters of the $\alpha$ and $\beta$ carbons in the headgroup are slightly larger in the ECC-POPC model than in the Lipid14 model, which is apparently related to the P-N vector orienting by about 7$^{\circ}$ more toward the water phase in the ECC-POPC model, see Fig.~\ref{OrderParameterCHANGESsurf}. Considering the available experimental evidence, it is not possible to suggest which of the two models provides more realistic headgroup orientation. The ECC-POPC model gives the $\beta$ carbon order parameter value closer to experiments than the Lipid14 model, while the opposite is true for the $\alpha$ carbon. The accuracy of both models in the glycerol backbone region is equally footed with any other state of art lipid model available in literature \cite{botan15}, see Fig.~\ref{simVSexpNOions}.

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    a molar fraction of the cationic surfactant dihexadecyldimethylammonium in a POPC bilayer
    from simulations and experiments \cite{scherer89} at 313 K.
  }
\end{figure}

Before studying the sodium and calcium ion binding affinities, we quantify the response of the headgroup order parameters to the amount of bound charge by using mixtures of monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89}. These mixtures have a well-defined amount of bound charge per PC, i.e., the molar fraction of cationic surfactants. This direct relation results from the ability of dihexadecyldimethylammonium to be inserted in the lipid bilayers as any lipid because of its two hydrophobic acyl chains. Furthermore, available experimental data for these systems can be used to validate the sensitivity of lipid headgroup order parameters to the amount of bound charge in simulations \cite{scherer89}.

The changes of the headgroup order parameters with increasing amount of the cationic surfactant are compared between simulations and experiments~\cite{scherer89} in Fig.~\ref{OrderParameterCHANGESsurf}. An approximately linear decrease of the order parameters, as expected from Eq. \ref{OPchangeEQ}, is observed in both simulations and experiments at least for mole fractions below $\sim$30\%. The slope is, however, too steep in the Lipid14 model indicating that the response of head group order parameters is too large to the bound positive charge. In contrast, the slope of the ECC-POPC model is in a very good agreement with experiments for the $\alpha$ segment, while being slightly underestimated for the $\beta$ segment.

In Fig.~\ref{OrderParameterCHANGESsurf}, we show the headgroup P-N vector angle as a function of the mole fraction of the cationic surfactant. As suggested previously~\cite{seelig87}, the headgroup orients more towards the water phase with the increasing amount of positive charge in a PC lipid bilayer. The effect is more pronounced in the Lipid14 model than in the ECC-POPC model.   For example, the addition of 50\% mole fraction of the cationic surfactant leads to the decrease of 20$^{\circ}$ of the P-N vector angle for the Lipid14 model while only 11$^{\circ}$ in the ECC-POPC model. The difference is in line with the smaller order parameter changes and the reduced charge--dipole interactions in the latter model. The lower sensitivity of the P-N vector angle response in the ECC-POPC model agrees better with experiments.
%The results imply that improvements are required in MD models for reliable studies of the lipid headgroup responses to ions or other biomolecules.
%{\color{red}., as also suggested previously \cite{botan15}} \todo{Do you want to finish each paragraph with a citation to your previous works?}.

>>>>

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89_POPC-struct.pdf}
  %\includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-POPC models compared with experiments~\cite{kucerka11} at 303~K. \\
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-POPC models
    compared with experiments \cite{ferreira13} at 300~K.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$
    (Bayesian estimate of 95\% confidence interval \cite{scipy}).
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Open/closed symbols are used for palmitoyl/oleoyl chains of POPC. \\
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
  } 
\end{figure}

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of head group order parameters and P-N vector orientation as a function of
    a molar fraction of the cationic surfactant dihexadecyldimethylammonium in a POPC bilayer
    from simulations and experiments \cite{scherer89} at 313 K.
  }
\end{figure}

@@
